Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-06 paper.tex
smc-description

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}

\title{bit-smc-paper}

@@

\linenumbers
\title{backwards-in-time SMC}

@@

% Middle Authors
\author[2]{Alison Etheridge}

\author[3]{Peter Ralph}
\author[1]{Jerome Kelleher}
\affil[$\dagger$]{Denotes corresponding author}

@@

\section{Abstract}

\section{Introduction}

\section{Methods}

\section{Results}

\section{Discussion}

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts, bbm}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{math,calc,positioning}
\usetikzlibrary{calc,positioning}
\usepackage{subcaption}

\newcommand{\noderef}[1]{\textsf{#1}}
\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}

@@

\linenumbers
\title{Computing likelihoods under the SMC for a general class of ARGs.}

@@

% Middle Authors
\author[2]{Alison Etheridge}
\author[3]{Peter Ralph}

% last author
\author[1]{Jerome Kelleher}
\affil[$\dagger$]{Denotes corresponding author}

@@

\section{Abstract}
%% WHAT IS THE MAIN FOCUS OF THE PAPER?

\section{Introduction}
% What are ARGs, their relevance and importance

% associating likelihoods with ARGs

% previous likelihood calculations on ARGsL Kuhner and Felsenstein (2000), ARGweaver

% history of approximating the coalescent: 
SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly 
(2000)
These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
Clear distinction into two groups: ARGWeaver infers actual ARG.
Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
structures. (other peculiarities: ...).

% SMC and relatives are useful: lot of focus however on integrating over the distribution of trees to get Ne(t). \cite{PSMC, ...}
% Here the idea is to use the SMC in a different way that resonates more with the true direction of flow
% of genetic/haplotype information.

% results in a concrete way to compute likelihoods for *any* inferred ARG (integrating out the recombination time)
% meaning: e.g. no requirements on adjacent trees only differing by a single SPR, (other?)

% In this paper, we attempt to address these problems by
% providing a precise definition of an ARG that is general enough to encompass
% the output of popular simulation and inference methods.

% ADD: remark on fact that we are only dealing with topology and branch length information
% adding likelihood of observing particular mutation configuration given an ARG is straightforward.

\section{Methods and Results}
\subsection{SMC backwards-in-time}\label{par:description}

Formulated backwards-in-time the SMC \citep{mcvean_approximating_2005} only requires a simple modification to the 
coalescent with recombination. By only considering those pairs of 
lineages that share 
genomic intervals with ancestral material, as eligible for coalescence, we obtain the 
sequential Markovian structure along the chromosome the SMC is best known for.\\

More formally, and following the notation of \cite{mcvean_approximating_2005}, at any 
point in time, the process can be described by $L(t)$, the set of labeled lineages 
extant at time $t$ each represented by a union of ordered non-overlapping half open 
intervals detailing the ancestral material 
carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by 
single interval spanning the entire genome.
Backwards in time the process evolves by a succession of coalescence and recombination 
events until all segments of ancestral material are each present in only one lineage. 
The waiting time to next event is determined by these two competing processes with exponentially 
distributed waiting times as described below.\\

Recombination is described by a Poisson process of rate $r$ per unit of length and time. 
A recombination to the 
left of $x$ with $x>x_{i0}$ and $x<y_{im}$ splits a lineage $i$ into two new, uniquely 
labelled lineages.\\

Any two \emph{overlapping} lineages coalesce at rate $\lambda = 1/(2N_e)$ in the case
of a diploid population of constant size. The newly formed lineage acquires the 
union of both ordered intervals.
Although coalescence is reciprocal, here we'll define a strict total order 
(see \ref{par:liks}) on 
all lineages in $X(t)$ based on their leftmost starting points. 
For any such strict total order on $X_l(t)$ and at any point in the process 
the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.

\begin{equation} \label{def:coal}
I_{ij} = \begin{cases}
1 & X_i > X_j \wedge X_i \cap X_j \neq \emptyset \\
0 & \text{otherwise}
\end{cases}
\end{equation}

We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
I_{i}(t) = \sum_{j} I_{ij}(t)$.
Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
(see next paragraph). Note that although recombination affects $C_i(t)$ it does not 
affect $I_{i}$.\\

% note on the disjoint set of intervals for each lineage under the SMC
As noted by \cite{mcvean_approximating_2005}, the backwards-in-time description of the 
SMC is slightly different from the left-to-right approach in that recombination 
in non-ancestral material is possible. Although this does not affect the distribution 
of marginal genealogies, it does mean that not all lineages in $L$ can be represented 
by a single interval. More precisely, all lineages will be associated with 
a single interval up to the point where one section of the 
genome has reached its most recent common ancestor but the neighbouring parts have not.

\subsection{Recording an ARG} \label{par:recording}

The process described in the previous paragraph details the transitions of the 
set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
$\mathcal{G} = (N, E)$ is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
replaced by one (coalescence) or two lineage(s) (recombination), $p$ (and $p^{\prime}$).
The last element of the tuple is the extent of the edge. It is itself a  
disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
and $N$, the set of nodes representing all genomes in the ARG, together
define this data structure.\\

Recording the full process as described above will result in unary nodes.
These are nodes that will have only one child in one or more local trees 
(see fig. \ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages hit by a recombination event. 
In case of a coalescence event, a node is registered along the full 
extent of the original lineage $X_c$. 
Coalescence between two lineages without perfectly 
overlapping intervals will therefore result in locally unary nodes.\\

The resulting data structure can be simplified to various degrees \citep{Wong-2023}. 
This equates to removing certain types of unary nodes, 
depending on the degree of information 
we want to retain (or can reconstruct) from the original process. 
Unary nodes therefore are a vital link between 
ARGs and local trees. In a fully simplified ARG, without unary nodes, 
the local trees only store information on local coalescence events.

Here we will compute the likelihood for an ARG
under the SMC for which we have no information on the time to any 
of the recombination events. We will assume however, that the identity 
of the lineage involved in each recombination event can be correctly 
inferred by relying on the unary nodes (see section X.X on locally unary nodes). 
This enables us to extend likelihood computations to the 
a more general class of ARGs.

\begin{figure}[h!] \label{fig:smc-unary}
\centering
\includegraphics[width=\textwidth]{smc-custom-haplotypes-mini.png}
\caption{SMC with unary nodes}
\end{figure}

\subsection{Unary nodes} \label{par:unary}

The SMC with unary nodes
Unary nodes connect marginal trees to the ARG.
These are inferred by \tsinfer, \kwarg, \argweaver. %add citation

% what happens in the partial or complete absence of unary nodes
% even in the absence of unary nodes we can compute a likelihood, however, this has an impact.

% The SMC and unary nodes
% ADD: figure here that tries to deal with all possible causes of confusion
% unary nodes mainly

% highlight the amount of information that needs to be tracked to compute a likelihood
% for the exact Hudson CwR

\subsection{Calculating likelihoods} \label{par:liks}

Retracing the ARG backwards in time, each coalescence event is associated with two child 
lineages, $c$ and $d$, and a single parent, $p$. By defining $C_i(t)$ asymmetrically, 
the history of each these two child lineages can be considered independently without 
risking to double count any of the coalescence events. The likelihood of 
observing the edge $(c, p, X_{cp})$ can thus be considered independently from 
the likelihood of all other edges, and in particular from the likelihood 
of $(d, p, X_{dp})$.\\

% extends from here means min(X_{cp})=x, max(X_{cp})=y
% improve this description here
Assuming $(c, p, X_{cp})$ extends from $x$ to $y$ along the genome and $t_i$ 
indicates the age of 
node $i$ in the ARG, then the probability of not observing a recombination across the 
entire area (span x depth) of this edge is

% full extent of segment
\begin{equation}\label{eq:span}
A_{(c, p)}(\theta) = e^{-r (y-x)(t_p - t_{c})}
\end{equation}
% also x and y not necessarily equal to x_{c_{1}0} and y_{c_{1}m}.

What's left to compute the likelihood, is to take into account all events 
associated with $x$, 
the leftmost coordinate of $(c, p, X_{cp})$. This is either only the 
coalescence event involving 
$c$ and $d$. In this case $x=x_{c0}$. Or, $x$ can a new recombination break point that 
occurred on the lineage represented by $c$ at some time $s$ between $t_{c}$ 
and $\min(t_p, t_{p^{\prime}})$, where $p^{\prime}$ is the parent of the 
edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$.
In the latter case the new lineage starting at $x$ can be given a temporary label 
$c^{\prime}$ such that with the same logic from the previous paragraph we 
can define $F(c^{\prime}, s, t) = \int_{s}^{t} I_{c^{\prime}}(u)du$.

\begin{equation}\label{eq:depth}
B_{(c, p)}(\theta) = \begin{cases}
e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c^{\prime}, s, t_{p})} ds \lambda^{I_{c^{\prime}d}(t_p)} & x=x_{c^{\prime}0}>x_{c0} \\
\end{cases}
\end{equation}

Here, the (second) exponential term gives the probability of not observing a 
coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further 
gives the probability of observing a recombination 
at time $s$ given a Poisson process along position $x$ with rate $r$. The 
time of the recombination is integrated out.
The last term gives the point probability density of the coalescence event 
that terminates the edge. The indicator function is used to resolve 
the non-reciprocal nature of coalescence events as defined here.\\

The likelihood of ARG $\mathcal{G}$ defined by the set of edges $E$ and
given parameters $\theta$ then is

\begin{equation}\label{eq:full-lik}
\mathcal{L}(\mathcal{G}|\theta) = \prod_{(c, p) \in E} A_{(c, p)}(\theta) * B_{(c, p)}(\theta)
\end{equation}

% Does this hold for the case beyond the root of the genealogy?

Because the likelihood is computed based on edges,
and edges are no longer recorded once the local most recent common ancestor 
has been reached, we will underestimate the number of positions where 
a recombination could have occurred in case a recombination hits a 
lineage with trapped non-ancestral material (see end \ref{par:description}). 
If we assume that only a single recombination 
occurred, then it suffices however to provide a correction factor $g$ in equation
\ref{eq:depth}. Here $g$ represents the 
length of non-ancestral material along which a recombination would have resulted  
in the same observed local sequence of tree changes.

\begin{figure}[h!] \label{fig:algo}
\centering
\includegraphics[width=0.9\textwidth]{smc_algo.png}
\caption{Description of algorithm}
\end{figure}

\begin{figure}[h!] \label{fig:algo-flat}
\centering
\includegraphics[width=\textwidth]{smc_algo_flat.png}
\caption{Description of algorithm}
\end{figure}

\subsection{Tree sequence format} \label{par:algo}

Although the argument outlined in the previous paragraph is valid for any ARG, here we 
will describe how to (efficiently) extract the information required to associate a
tree sequence encoding of an ARG with a likelihood under the SMC.\\

Formally, we represent a tree sequence using a set of tables \citep{kelleher_efficient_2018}. 
This simple tabular representation of any ARG $\mathcal{G} = (N, E)$, can be 
augmented with additional tables with information on sequence variation
using a site and mutation table. The node table stores information on all 
haplotypes $N$ in the tree sequence. Information about how nodes relate to 
each other along the genome is defined in a tabular encoding of $E$.
Each row in this edge table constitutes a tuple \texttt{(left, right, parent, child)}.
Note that we defined the extent of any edge $\epsilon \in \mathcal{G}$ as the union of 
disjoint intervals. In the tree sequence encoding however, $\epsilon$ will be represented 
by a single entry for each contiguous interval. As remarked before, under the SMC, this 
distinction only matters beyond the root of the genealogy. The distinction between both 
types of edges will therefore not be made in the remainder of the paper.\\

The likelihood computation as described above can be done with a single pass 
of the edge table.
The central problem however is to efficiently maintain the state $I_e(t)$, which counts 
the number of lineages (the lineage associated with) the focal edge $\epsilon$ can coalesce 
with at time $t$.

The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
that we can impose a strict total order on all edges based on their starting point.

Because of the SMC:
an edge starting at $x$ can only coalesce 
with a subset of the other edges that make up the local tree at position $x$.

Because local tree changes only occur whenever edges are removed and/or inserted,
we will rely on the ordering of these insertions and removals as implemented in
\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\

% edge diff algorithm
The order in which edges are inserted and removed is determined by the index 
vectors $\textbf{i}$ and $\textbf{o}$ \citep{kelleher_efficient_2016}. The edge insertion vector 
$\textbf{i}$ gives the ordering of edges sorted by left endpoint, and among edges 
with the same left endpoint, sorted so that edges closer to the root appear later. 
The edge removal vector $\textbf{o}$ is similar, except gives the ordering of edges 
by right endpoint and with edges closer to the root appearing sooner. This order imposes the 
strong ordering on all edges as required by the definition of $I_{kl}$. The variables $\texttt{i}$ and 
$\texttt{o}$ variables are used to keep track of our position in the edge insertion and edge removal 
indexes.\\

% the counting algorithm
To transition from one tree to the next, we first iterate across the indices of all edges 
in $\textbf{o}$ from its current index up to $\texttt{o}$. For each edge in we decrement $I$ 
from index \texttt{edge.child} to \texttt{edge.parent}. We subsequently iterate across the 
edges in $\textbf{i}$ up to index $\texttt{i}$. For each new edge, we first compute the likelihood 
associated with that edge and then 
increment $I$ from \texttt{edge.child} to \texttt{edge.parent}.\\

% improve this section, then transition to limitations of unary nodes
We further rely on the assumption that the (inferrable, see next paragraph) unary nodes 
provide us with the required information on both the extent of all extant lineages through 
time as well as on the time boundaries for any recombination event and the identity of the 
lineage hit by that recombination.

% section on how deviations from the ideal assumptions affect the algorithm
% extent of lineages
% version of the algorithm with polytomies

% version of the algorithm in the absence of unary nodes

\subsection{Flexibility}

By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence.
Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
benefit from this. %elaborate
We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).
Similarly, any subset of the tree sequence that is itself a valid tree sequence can again 
be used to compute a likelihood. Such operations include: slicing genome both in time, and space.

\subsubsection{Slicing the ARG in time}
% TO DO: give example why this might be useful.

\subsubsection{Slicing the ARG in (genome) space}
%TO DO: give example why this might be useful.

\subsubsection{Variable Ne}

\subsubsection{Variable r}

%\subsubsection{Node labellings}
% implement idea from Guo et al. 2022.

\section{Discussion}
% summary: what did we do here

% future directions: 
incorporating node labels (see \citep{guo_recombination-aware_2022}).

First step towards a genealogy sampler (for any ARG, and at scale): quantifying uncertainty 
(with inferred ARG as starting point)

resolve polytomies in a model-based way

% explore the information in unary nodes

\bibliographystyle{plainnat}
\bibliography{paper, temp}

\section{Supplementary material}
